---
title: "Penalized B-spline"
author: "Gayoung Moon"
date: "2025-06-06"
institute: School of Mathematics, Statistics and Data Science \newline Sungshin Women’s University
fonttheme: "serif"
fontsize: 8pt
output:
  beamer_presentation:
    latex_engine: xelatex 
    theme: "metropolis"
header-includes:
  - \input{header_includes.tex}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

# Introduction

## What Is Penalized B-spline

- Penalized B-spline(P-spline) is a spline method that adds a penalty term to B-spline to create a smoother curve.
\vt
\vt

- P-spline can solve the problems that occur in B-spline regression:
\vt
  - The problem sensitive to number or location of knots.
\vt
  - The problem that the model is too flexible and vibrates.


## The Function of P-spline

- The following is the function of P-spline:
$$
\hat{y}= \sum^K_{j=1} \beta_j B_j(x)
$$
\vt

and we minimize the equation that follows as:
$$
\begin{aligned}
&\min_\beta \{ \Vert y - B \beta \Vert^2 + \lambda \Vert D^{(d)} \beta \Vert^2 \} \\
= &\min_\beta \bigg \{\sum^n_{i=1} \Big(y_i - \sum^K_{j=1} \beta_j B_j(x_i) \Big)^2 + \lambda \sum^K_{j=1}( \Delta^d \beta_j)^2 \bigg \}.
\end{aligned}
$$


## The Coefficient of P-spline

- We can also compute the coefficient of P-spline $\hat{\beta}$ like:
$$
\hat{\beta} = \Big( B^{\top}B + \lambda D^{(d)\top}D^{(d)} \Big)^{-1} B^{\top} y.
$$
\vt
\vt

- $y$: vector of the responses that are observed,
\newline
$B$: design matrix based on B-spline basis,
\newline
$\beta$: vector of B-spline coefficients,
\newline
$\lambda$: smoothing parameter,
\newline
$D^{(d)}$: difference matrix that computes the $d$-th order difference of $\beta$.


# **P-spline** Package in R

## P-spline Package

- P-spline package is a package that receives the number of knots, adjusts their positions, and performs penalized b-spline regression.
\vt
\vt

```{r, results= 'asis', echo= FALSE}
knitr::kable(data.frame(
  'function' = c("knots", "bspline_basis", "fit_pspline", "predict_pspline", "plot_pspline"),
  'explain' = c("Function of knots sequence using the quantiles of the data", "B-spline basis function and design matrix of B-spline basis", "Fitting P-spline model", "Predicting the values of y for given x based on P-spline", "Drawing a plot of P-spline")
), align = c('c', 'l'))
```


## The Variables Using in Package

```{r, results= 'asis', echo= FALSE}
knitr::kable(data.frame(
  'variable' = c("x", "new_x", "grid_x", "x_values", "y_values", "model", "interior_knots", "dimension", "degree", "lambda", "diff_order"),
  'explain' = c("Numeric vector of the data", "Numeric vector expressing a grid of evaluation points", "Numeric vector with a grid of evaluation points", "Numeric vector of x (input)", "Numeric vector of y (response)", "Function that fit P-spline with given data points",  "Vector of interior knots", "Number of the basis function", "Degree of spline(default= 3)", "Smoothing parameter (default = 1e-2)", "Order of the difference penalty (default = 2)")
), align = c('c', 'l'))
```


## Computing Knots Using **quantile**

```{r, echo= TRUE}
knots_quantile <- function(x, dimension, degree = 3){
  dimension = max(dimension, degree + 1)
  num_interior = dimension - degree - 1
  
  if (num_interior > 0){
    probs <- (1:num_interior)/(num_interior + 1)
    interior_knots <- quantile(x, probs, type = 1)
  } 
  else{
    interior_knots <- numeric(0)
  }

  return(interior_knots)
}
```


## Adding Boundary Knots to Interior Knots

```{r, echo= TRUE}
# Add boundary knots to interior knots.

add_boundary_knots <- function(x, interior_knots, degree = 3, tiny = 1e-5){
  knots<- c(rep(min(x) - tiny, degree + 1),
            interior_knots, rep(max(x) + tiny,
            degree + 1))
  return(knots)
}
```


## Definition B-spline Basis Function

```{r, echo= TRUE}
# B-spline basis function

b_spline <- function(x, knots, degree, i){
  # Base case: 0th degree
  if (degree == 0){
    return(ifelse(knots[i] <= x & x < knots[i + 1], 1, 0))
  }

  # Recursive case: degree > 0
  B_i_d1 <- b_spline(x, knots, degree - 1, i)
  B_i1_d1 <- b_spline(x, knots, degree - 1, i + 1)

  denom1 <- knots[i + degree] - knots[i]
  denom2 <- knots[i + degree + 1] - knots[i + 1]

  term1 <- if (denom1 == 0) 0 else ((x - knots[i])/denom1)*B_i_d1
  term2 <- if (denom2 == 0) 0 else ((knots[i + degree + 1] - x)/denom2)*B_i1_d1

  return(term1 + term2)
}
```


## Definition Design Matrix of B-spline Basis

```{r, echo= TRUE}
# Create design matrix of B-spline basis

create_design_matrix <- function(x_values, knots, degree){
  n <- length(x_values) # number of the data
  num_basis <- length(knots) - degree - 1 # number of the basis function

  design_matrix <- matrix(0, nrow = n, ncol = num_basis)

  for (j in 1:num_basis){
    for (i in 1:n){
      design_matrix[i, j] <- b_spline(x_values[i], knots, degree, j)
    }
  }

  return(design_matrix)
}
```


## Fitting P-spline Model

```{r, echo= TRUE}
fit_pspline <- function(x_values, y_values,
                        interior_knots, degree = 3,
                        lambda = 0.01, diff_order = 2){

  knots <- add_boundary_knots(x_values, interior_knots, degree)

  G <- create_design_matrix(x_values, knots, degree)
  # Design Matrix of B-spline

  num_basis <- ncol(G)
  D <- diff(diag(num_basis), differences = diff_order)

  P <- t(D) %*% D # Penalty Matrix

  beta_hat <- solve(t(G) %*% G + lambda * P) %*% t(G) %*% y_values

  return(list(beta = beta_hat, knots = knots,
              degree = degree, lambda = lambda,
              penalty = P))
}
```


## Predicting the Values of Data Based on P-spline

```{r, echo= TRUE}
predict_pspline <- function(model, new_x){
  G_new = create_design_matrix(new_x, model$knots, model$degree)
  y_pred = G_new %*% model$beta
  return(y_pred)
}
```


## Plotting P-spline with Scatter Plots

```{r, echo= TRUE}
library(ggplot2)

plot_pspline <- function(x_values, y_values, model, grid_x, num_knots, lambda,
        point_size, line_size){
  y_pred = predict_pspline(model, grid_x)
  data_plot = data.frame(x = x_values, y = y_values)
  spline_plot = data.frame(x = grid_x, y = y_pred)

  ggplot() +
    geom_point(data = data_plot, aes(x, y),
               color = "blue", size= point_size) +
    geom_line(data = spline_plot, aes(x, y),
              color = "red", linewidth = line_size) +
    labs(title =
        sprintf("Fitted P-spline Regression (knots= %d, lambda= %.2f)",
        num_knots, lambda), x = "x", y = "y") +
    xlim(c(min(x_values), max(x_values)))
}
```


# Apply Data set

## Generating Data set

```{r, echo= TRUE}
set.seed(924)
n= 100
x_values= sort(runif(n, 0, 1))
y_values= sin(2*pi*x_values) + cos(4*pi*x_values) + rnorm(n, sd= 0.2)
```


## Using **P-spline** Package

```{r, echo= TRUE, eval= FALSE}
# devtools::install_github('Ga-young-Moon/Penalized-B-spline')
library(Pspline)

# spline degree specification
degree= 3

# knot generation
num_interior= 5
interior_knots= knots_quantile(x_values, num_interior)

# setting the value of lambda
# and the difference penalty order
lambda= 1e-2
diff_order= 2

# model fitting
model= fit_pspline(x_values, y_values, interior_knots,
                   degree, lambda, diff_order)
```

```{r, echo= FALSE, eval= TRUE}
# devtools::install_github('Ga-young-Moon/Penalized-B-spline')
library(Pspline)

# spline degree specification
degree= 3

# knot generation
num_interior= 5
interior_knots= knots_quantile(x_values, num_interior)

# setting the value of lambda
# and the difference penalty order
lambda= 1e-2
diff_order= 2

# model fitting
model= fit_pspline(x_values, y_values, interior_knots,
                   degree, lambda, diff_order)
```


## Drawing the Plot of Data Point

```{r, echo= TRUE, out.width= '87%', fig.align='center'}
grid_x = seq(min(x_values), max(x_values), length.out = 100)
plot_pspline(x_values, y_values, model, grid_x, 5, 1e-2, 1.5, 0.5)
```


## Plot Shape According to Number of Knots (Knots= 7)

- From the plot above, we can see that when there are 5 knots, the spline curve does not follow the data very well. \newline
$\rightarrow$ Therefore, we need to check the plot when the number of knots is increased.

- If the number of knots is 7:

```{r, echo= FALSE, eval= TRUE, , out.width= '87%', fig.align='center'}
num_interior= 7
interior_knots= knots_quantile(x_values, num_interior)
lambda= 1e-2
diff_order= 2
model= fit_pspline(x_values, y_values, interior_knots,
                   degree, lambda, diff_order)

grid_x = seq(min(x_values), max(x_values), length.out = 100)
plot_pspline(x_values, y_values, model, grid_x, num_interior, lambda, 1.5, 0.5)
```


## Plot Shape According to Number of Knots (Knots= 15)

- If the number of knots is 15:

```{r, echo= FALSE, eval= TRUE, out.width= '87%', fig.align='center'}
num_interior= 15
interior_knots= knots_quantile(x_values, num_interior)
lambda= 1e-2
diff_order= 2
model= fit_pspline(x_values, y_values, interior_knots,
                   degree, lambda, diff_order)

grid_x = seq(min(x_values), max(x_values), length.out = 100)
plot_pspline(x_values, y_values, model, grid_x, num_interior, lambda, 1.5, 0.5)
```


## Plot Shape According to Number of Knots (Knots= 25)

- If the number of knots is 25:

```{r, echo= FALSE, eval= TRUE, out.width= '87%', fig.align='center'}
num_interior= 25
interior_knots= knots_quantile(x_values, num_interior)
lambda= 1e-2
diff_order= 2
model= fit_pspline(x_values, y_values, interior_knots,
                   degree, lambda, diff_order)

grid_x = seq(min(x_values), max(x_values), length.out = 100)
plot_pspline(x_values, y_values, model, grid_x, num_interior, lambda, 1.5, 0.5)
```

- As the number of knots increases, we can see that the spline model becomes more flexible.


## Plot Shape According to Value of $\lambda$ ($\lambda$ = 0.01)

- Then, we can also see how the plot changes depending on $\lambda$ value.

```{r, echo= FALSE, eval= TRUE, out.width= '87%', fig.align='center'}
num_interior= 25
interior_knots= knots_quantile(x_values, num_interior)
lambda= 1e-2
diff_order= 2
model= fit_pspline(x_values, y_values, interior_knots,
                   degree, lambda, diff_order)

grid_x = seq(min(x_values), max(x_values), length.out = 100)
plot_pspline(x_values, y_values, model, grid_x, num_interior, lambda, 1.5, 0.5)
```


## Plot Shape According to Value of $\lambda$ ($\lambda$ = 1)

- If $\lambda$= 1:

```{r, echo= FALSE, eval= TRUE, out.width= '87%', fig.align='center'}
num_interior= 25
interior_knots= knots_quantile(x_values, num_interior)
lambda= 1
diff_order= 2
model= fit_pspline(x_values, y_values, interior_knots,
                   degree, lambda, diff_order)

grid_x = seq(min(x_values), max(x_values), length.out = 100)
plot_pspline(x_values, y_values, model, grid_x, num_interior, lambda, 1.5, 0.5)
```


## Plot Shape According to Value of $\lambda$ ($\lambda$ = 0.0001)

- If $\lambda$= 0.0001:

```{r, echo= FALSE, eval= TRUE, out.width= '87%', fig.align='center'}
num_interior= 25
interior_knots= knots_quantile(x_values, num_interior)
lambda= 1e-4
diff_order= 2
model= fit_pspline(x_values, y_values, interior_knots,
                   degree, lambda, diff_order)

grid_x = seq(min(x_values), max(x_values), length.out = 100)
plot_pspline(x_values, y_values, model, grid_x, num_interior, lambda, 1.5, 0.5)
```

- The smaller $\lambda$ value, the better the curve follows the data.


## Q & A

\begin{center}
  {\bf {\Huge Q \& A}}
\end{center}


## 

\begin{center}
  {\bf {\Huge Thank you :)}}
\end{center}
